---
title: "STA141C Final Project Code"
author: "Matthew Holcomb"
date: "2024-06-10"
output: html_document
---

This file contains the R Code used in our final project. It is a combination of
two different files for both the Pre-Processing and Logistic Regression portions.

```{r Get Data}
# Read in the full data set
NBA_Data <- read.csv("NBA_Player_Stats.csv")
NBA_Data <- NBA_Data[,-1] # Remove an unnecessary column which R generates

NBA_Data$All_NBA <- as.logical(NBA_Data$All_NBA)

rownames(NBA_Data) <- paste(NBA_Data$Player, NBA_Data$Year, NBA_Data$Tm)

columns <- c("Year", "Pos", "All_NBA", "Voting_Points", "All_NBA_Team", 
             "PTS", "TRB", "AST", "BLK", "STL", "W.L", "FGA", "FG.", "eFG.", 
             "FT", "FT.", "MP", "G", "GS")

stat_columns = c("PTS", "TRB", "AST", "BLK", "STL", "W.L", "FGA", 
                 "FG.", "eFG.", "FT", "FT.", "MP", "G", "GS")

data = NBA_Data[,columns]

# We will only standardize the columns that we will be using as our predictors
data.scaled <- data

# Standardize the data by season using scale()
for (year in 2004:2024) {
  data.scaled[data.scaled$Year == year,stat_columns] <- scale(data[data$Year == year,stat_columns])
}

# write.csv(data.scaled,"Data/Scaled_Data.csv", row.names = TRUE)
# write.csv(data, "Data/Organized_Data.csv", row.names = TRUE)
```

\newpage

# Visualizing the Transformed Data

We can visualize the transformed data using the `PTS` values for three different years.
The mean is listed next to the year in the plot titles.

```{r, echo=F, fig.dim=c(6,4)}
par(mfrow = c(2,3))

# Original (2004, 2012, 2023)
hist(data[data$Year == 2004,]$PTS, 
     main = paste0("Original (2004, ", 
                  round(mean(data$PTS[data$Year == 2004]), 3),
                  ")"),
     xlab = "Original PTS", ylab = "Frequency", col = "lightblue")
hist(data[data$Year == 2012,]$PTS, 
     main = paste0("Original (2012, ", 
                  round(mean(data$PTS[data$Year == 2012]), 3),
                  ")"),
     xlab = "Original PTS", ylab = "Frequency", col = "lightblue")
hist(data[data$Year == 2023,]$PTS, 
     main = paste0("Original (2023, ", 
                  round(mean(data$PTS[data$Year == 2023]), 3),
                  ")"),
     xlab = "Original PTS", ylab = "Frequency", col = "lightblue")

# Transformed (2004, 2012, 2023)
hist(data.scaled[data.scaled$Year == 2004,]$PTS, 
     main = paste0("Transformed (2004, ", 
                  round(mean(data.scaled$PTS[data.scaled$Year == 2004]), 3),
                  ")"),
     xlab = "Transformed PTS", ylab = "Frequency", col = "lightblue")
hist(data.scaled[data.scaled$Year == 2012,]$PTS, 
     main = paste0("Transformed (2012, ", 
                  round(mean(data.scaled$PTS[data.scaled$Year == 2012]), 3),
                  ")"),
     xlab = "Transformed PTS", ylab = "Frequency", col = "lightblue")
hist(data.scaled[data.scaled$Year == 2023,]$PTS, 
     main = paste0("Transformed (2023, ", 
                  round(mean(data.scaled$PTS[data.scaled$Year == 2023]), 3),
                  ")"),
     xlab = "Transformed PTS", ylab = "Frequency", col = "lightblue")
```

The transformation allows each year to theoretically be compared evenly. For instance,
if a player in 2004 scored 500 total points and was the top scorer for that year, but
in 2023, ten players scored 500+ points, we would want the model to recognize the
500 point season in 2004 as more influential/impressive.

## Tables for Presentation Slides

```{r}
library(knitr)
tbl_columns = c("Year", "All_NBA", "PTS", "AST", "TRB", "W.L")
tbl_players = c("LeBron James 2023 LAL", "Stephen Curry 2023 GSW", "De'Aaron Fox 2023 SAC")
kable(data[tbl_players,tbl_columns])

kable(data.scaled[tbl_players,tbl_columns])
```

```{r}
Scaled_vs_Regular <- rbind(
  data["LeBron James 2021 LAL",tbl_columns],
  data["LeBron James 2022 LAL",tbl_columns],
  data.scaled["LeBron James 2021 LAL",tbl_columns],
  data.scaled["LeBron James 2022 LAL",tbl_columns]
)

Scaled_vs_Regular[,3:6] <- round(Scaled_vs_Regular[,3:6], 3)
rownames(Scaled_vs_Regular) <- c(
  "LeBron James 2021 LAL",
  "LeBron James 2022 LAL",
  "LeBron James 2021 LAL (Scaled)",
  "LeBron James 2022 LAL (Scaled)"
)

kable(Scaled_vs_Regular)
```

\newpage

# Data Exploration

```{r}
library(ggplot2)

ggplot() +
  geom_point(data = subset(data.scaled, Year < 2024 & !All_NBA), 
             mapping = aes(x = PTS, y = AST, color = "FALSE")) +
  geom_point(data = subset(data.scaled, Year < 2024 & All_NBA), 
             mapping = aes(x = PTS, y = AST, color = "TRUE")) +
  labs(title = NULL,
       x = "Points (Scaled)", y = "Assists (Scaled)",
       color = "All-NBA") +
  theme(text = element_text(size = 20))

ggsave(
  filename = "PTS_AST.png",
  device = "png"
)

ggplot(data = subset(data.scaled, Year < 2024)) +
  geom_boxplot(mapping = aes(x = W.L, color = All_NBA)) +
  labs(title = NULL,
       x = "W/L Percentage (Scaled)", y = "",
       color = "All-NBA") +
  theme(
    text = element_text(size = 20),
    axis.text.y = element_blank()
  )
  

ggsave(
  filename = "W_L.png",
  device = "png"
)
```

```{r Get Data, echo = F}
# data.scaled <- read.csv("Data/Scaled_Data.csv", row.names = 1)

data.scaled$Voting_Points <- as.integer(data.scaled$Voting_Points)
```

### Logistic Regression Results Plot

```{r LogReg and Plot, echo = F, warning = F}
library(tidyr)
library(MASS)
library(knitr)

Do_LogReg <- function(the.data, year, year_range = 5, Positionless = FALSE, Cutoff = 0, Full = FALSE) {

  # Identify the training and test data
  train <- subset(the.data, Year %in% (year-year_range):(year-1))
  test <- subset(the.data, Year == year & G > Cutoff)

  # Initialize the full model
  full.model <- glm(All_NBA ~ . -Voting_Points -All_NBA_Team -Year, 
                    data = train, family = binomial)
  
  # Find the best model using stepAIC
  step.model <- full.model %>% stepAIC(trace = FALSE, direction = "backward")
  
  # Store the coefficients used in the best model
  coefficients <- coef(step.model)
  
  # Make predictions using new model
  test$prob <- predict(step.model, newdata = test, type = "response")
  
  predicted <- subset(test[order(test$prob, decreasing = TRUE),], 
                      select = c("Pos", "prob", "All_NBA"))
  
  if (Full) {
    new_predicted <- predicted
  } else {
    if (Positionless) {
      new_predicted <- predicted[1:15,]
    } else {
      FC <- subset(predicted, Pos == "Frontcourt")
      BC <- subset(predicted, Pos == "Backcourt")
      
      new_predicted <- data.frame()
      
      for (i in 0:2) {
        new_predicted <- rbind(new_predicted, BC[(1 + i*2):((i+1)*2),])
        new_predicted <- rbind(new_predicted, FC[(1 + i*3):((i+1)*3),])
      }
    }
    
    new_predicted$Pred_Team <- c(rep(1,5), rep(2,5), rep(3,5))
  }
  
  return(list(new_predicted = new_predicted, coefficients = coefficients))
}

Results_Two <- list()
Results_Four <- list()
Results_Eight <- list()

Coefficients_Two <- list()
Coefficients_Four <- list()
Coefficients_Eight <- list()

YEARS <- 2012:2023

for (year in YEARS) {
  two_years <- Do_LogReg(data.scaled, year, year_range = 2)
  four_years <- Do_LogReg(data.scaled, year, year_range = 4)
  eight_years <- Do_LogReg(data.scaled, year, year_range = 8)
  
  Results_Two[[as.character(year)]] <- two_years$new_predicted
  Coefficients_Two[[as.character(year)]] <- two_years$coefficients
  
  Results_Four[[as.character(year)]] <- four_years$new_predicted
  Coefficients_Four[[as.character(year)]] <- four_years$coefficients
  
  Results_Eight[[as.character(year)]] <- eight_years$new_predicted
  Coefficients_Eight[[as.character(year)]] <- eight_years$coefficients
}

Accuracies <- data.frame()
Snubs <- c()

for (year in YEARS) {
  correct_two <- sum(Results_Two[[as.character(year)]]$All_NBA)
  correct_four <- sum(Results_Four[[as.character(year)]]$All_NBA)
  correct_eight <- sum(Results_Eight[[as.character(year)]]$All_NBA)
  
  Accuracies <- rbind(Accuracies, data.frame(correct_two, correct_four, correct_eight, year))
}

ggplot(data = Accuracies) +
  geom_line(aes(x = year, y = correct_two, color = "2")) +
  geom_line(aes(x = year, y = correct_four, color = "4")) +
  geom_line(aes(x = year, y = correct_eight, color = "8")) +
  labs(title = "All-NBA Prediction Accuracy by Year",
       x = "Year", y = "# Correct",
       color = "n") +
  ylim(0, 15)

ggsave(
  filename = "LogReg.png",
  device = "png",
  width = 10,
  height = 6,
  units = 'cm'
)
```

\newpage

### Coefficents Table

The following table contains coefficients chosen through backward step-wise selection,
where the training data used was the previous 4 years to the listed season.

```{r Coefs, echo = F}
stat_columns <- colnames(data.scaled)[6:(length(data.scaled))]
stat_columns <- c(stat_columns, c("PosFrontcourt", "PosBackcourt"))

M = matrix(data = 0, nrow = length(YEARS), ncol = length(stat_columns)+1,
           dimnames = list(YEARS, c("(Intercept)", stat_columns)))

for (year in as.character(YEARS)) {
  df <- data.frame(
    name = names(Coefficients_Four[[year]]),
    val = round(as.numeric(Coefficients_Four[[year]]), 2)
  )
  for (i in 1:nrow(df)) {
    M[year, df$name[i]] <- df$val[i]
  }
}

print(M)
```

\newpage

### Snubs and Reaches

```{r Snubs, echo = F}
Snubs <- c()
Years <- c()

for (year in YEARS) {
  snubs_scaled <- rownames(subset(Results_Four[[as.character(year)]], !All_NBA))
  
  Snubs <- c(Snubs, snubs_scaled)
  Years <- c(Years, rep(as.character(year), length(snubs_scaled)))
}

Snubs <- sapply(Snubs, FUN = function(x) return(unlist(strsplit(x, " 2"))[1]), simplify = FALSE)
Snubs <- unlist(Snubs)

Snubs_df <- data.frame(
  table(Snubs)
)

Snubs_df$Years <- sapply(X = Snubs_df$Snubs, FUN = function(x) paste(Years[Snubs == x], collapse=" "))

Snubs_df <- Snubs_df[order(Snubs_df$Freq, decreasing = TRUE),]
rownames(Snubs_df) <- NULL

kable(subset(Snubs_df, Freq > 1))
```

```{r Reaches, echo = F}
Reaches <- c()
Years <- c()

for (year in YEARS) {
  actual <- rownames(subset(data.scaled, Year == year & All_NBA))
  pred_scaled <- rownames(Results_Four[[as.character(year)]])
  
  reaches_scaled <- actual[!(actual %in% pred_scaled)]
  
  Reaches <- c(Reaches, reaches_scaled)
  Years <- c(Years, rep(as.character(year), length(reaches_scaled)))
}

Reaches <- sapply(Reaches, FUN = function(x) return(unlist(strsplit(x, " 2"))[1]), simplify = FALSE)
Reaches <- unlist(Reaches)

Reaches_df <- data.frame(
  table(Reaches)
)

Reaches_df$Years <- sapply(X = Reaches_df$Reaches, FUN = function(x) paste(Years[Reaches == x], collapse=" "))

Reaches_df <- Reaches_df[order(Reaches_df$Freq, decreasing = TRUE),]
rownames(Reaches_df) <- NULL

kable(subset(Reaches_df, Freq > 1))
```

\newpage

### 2024 Predictions

```{r Pred_Table, echo = F, warning = F}
kable(subset(Do_LogReg(data.scaled, 2024, Cutoff = 0.738, Full = FALSE,
                       Positionless = TRUE, year_range = 4)$new_predicted, 
             select = prob))
```

```{r}
for (year in YEARS) {
  n = nrow(subset(data.scaled, Year == year))
  incorrect = 15 - Accuracies$correct_four[Accuracies$year == year]
  print((n - incorrect) / n)
}
```
